---
title: 'Data Visualization with ggplot2 in R'
subtitle: 'Activity #2: Customizing Themes'
format: html 
editor: visual
toc: true
---

## 0.0 Getting Started

Load the appropriate packages. Recreate the base plot you generated in Activity #1. Store the plot as `plot`.

```{r}
#| warning: false
library(ggplot2)
library(palmerpenguins)
```

## 1.0 Using pre-packaged themes

Add a pre-packaged theme to your base plot:

-   theme_grey()
-   theme_bw()
-   theme_linedraw()
-   theme_light()
-   theme_dark()
-   theme_minimal()
-   theme_classic()

## 2.0 Customizing your own theme

### 2.1 Add fonts

Add fonts that R can use for customizing plots.

```{r}
#| eval: false

#### For Windows Users ####
install.packages('extrafont')
library('extrafont')

font_import()
# You will be prompted to continue [y/n]
# Type 'y' and press enter

loadfonts(device="win")


#### For Mac Users ####
install.packages('extrafont')
library('extrafont')

font_import()
# You will be prompted to continue [y/n]
# Type 'y' and press enter

loadfonts()
```

Check what fonts are available with `fonts()`.

### 2.2 Customizing theme elements

Using the `theme()` function, customize the following theme elements:

-   plot.title
-   plot.subtitle
-   panel.background
-   panel.border
-   panel.background
-   panel.grid
-   axis.title
-   axis.text

You should use some combination of the following adjustments in your theme:

-   color
-   fill
-   size
-   linewidth
-   family
-   face (bold, italic, etc.)

**Be creative**. Effective data communication requires important information to be interpretable *and* visually inviting.

### 2.3 Saving your theme

Store your theme as `my_theme`.

### 2.4 Reusing your theme

Run the following code:

```{r}
#| eval: false
plot + my_theme
```

### 3.0 Examples

```{r}
#| echo: false
#| warning: false
#| fig-height: 6.5
library(extrafont)
ggplot(penguins) + 
  geom_point(aes(body_mass_g, flipper_length_mm, color = species)) + 
  labs(title = 'Flipper Length by Body Mass in Penguins',
       subtitle = 'Species type in color', 
       x = 'Body mass (g)',
       y = 'Flipper length (mm)', 
       color = 'Species') + 
  scale_color_manual(values = c('royalblue','orange','grey30')) + 
  theme(plot.title = element_text(color='cornflowerblue', size = 15, family = 'Comic Sans MS',face = 'bold'), 
        plot.subtitle = element_text(color = 'grey40', size = 12, family = 'Malgun Gothic'), 
        panel.background = element_rect(fill = 'whitesmoke'), 
        panel.grid = element_line(color=NA), 
        axis.title = element_text(color = 'grey20', size = 12, family = 'Malgun Gothic'),
        axis.text = element_text(color = 'grey40', size = 8, family = 'Malgun Gothic'), 
        legend.title = element_text(color = 'grey20', size = 12, family = 'Malgun Gothic'),
        legend.text = element_text(color = 'grey40', size = 8, family = 'Malgun Gothic'), 
        legend.position = 'bottom', 
        legend.key = element_rect(fill = NA)
        ) + 
  guides(color = guide_legend(override.aes = list(size = 3)))
```
