---
title: 'Data Visualization with ggplot2 in R'
subtitle: 'Activity #1: Creating Base Plots'
format: html 
editor: visual
---

## 0.0 Getting Started

Install and load the **palmerpenguins** package. You should notice that you now have access to the `penguins` dataset, even though it does not appear in the global environment.

```{r}
#| eval: false
install.packages("palmerpenguins")
```

```{r}
#| warning: false
library(ggplot2)
library(palmerpenguins)
head(penguins)
```

\newpage

## 1.0 Creating a Base Plot

Using the `penguins` dataset, create a scatterplot (using `geom_point()`) of flipper_length_mm (Y axis) by body_mass_g (X axis). Your plot should include *each* of the following:

-   Title
-   Subtitle
-   X-axis title
-   Y-axis title

*NOTE: Reference the code available to you in the slides or the code-along .R file.*

## 2.0 Adding a Layer

Add a layer to your base plot with any *one* of the following:

-   Mapped aesthetic
-   Scale change (axis, color, size, etc.)
-   Facet

\newpage

## 3.0 Examples

```{r}
#| echo: false
#| warning: false
#| fig-height: 3.5
ggplot(penguins) + 
  geom_point(aes(body_mass_g, flipper_length_mm, color = species)) + 
  labs(title = 'Flipper Length by Body Mass in Penguins',
       subtitle = 'Species type in color', 
       x = 'Body mass (g)',
       y = 'Flipper length (mm)', 
       color = 'Species')
```

```{r}
#| echo: false
#| warning: false
#| fig-height: 3.5
ggplot(penguins) + 
  geom_point(aes(body_mass_g, flipper_length_mm)) + 
  scale_x_continuous(limits = c(0,6500)) + 
  scale_y_continuous(limits = c(0,240)) + 
  labs(title = 'Flipper Length by Body Mass in Penguins',
       subtitle = 'Rescaled to x=0, y=0', 
       x = 'Body mass (g)',
       y = 'Flipper length (mm)')
```

```{r}
#| echo: false
#| warning: false
#| fig-height: 3.5
ggplot(penguins) + 
  geom_point(aes(body_mass_g, flipper_length_mm)) +
  facet_wrap(~species) + 
  labs(title = 'Flipper Length by Body Mass in Penguins',
       subtitle = 'faceted by species', 
       x = 'Body mass (g)',
       y = 'Flipper length (mm)')
```
